---
title: "Chapter 5"
author: "weh"
date: '2022-04-27'
output: html_document
---
>data

```{r}
# load data and copy 
library(rethinking) 
data(WaffleDivorce) 
d <- WaffleDivorce
# standardize variables
d$D <- standardize( d$Divorce ) 
d$M<- standardize( d$Marriage )
d$A <- standardize( d$MedianAgeMarriage )
```

>Divorce rate is associated with median age at marriage  m5.1

```{r}
m5.1 <- quap(
    alist(
        D ~ dnorm( mu , sigma ) ,
        mu <- a + bA * A ,
        a ~ dnorm( 0 , 0.2 ) ,
        bA ~ dnorm( 0 , 0.5 ) ,
        sigma ~ dexp( 1 )
    ) , data = d )

# compute percentile interval of mean
A_seq <- seq( from=-3 , to=3.2 , length.out=30 ) 
mu <- link( m5.1 , data=list(A=A_seq) )
mu.mean <- apply( mu , 2, mean ) 
mu.PI <- apply( mu , 2 , PI )
# plot it all
plot( D ~ A , data=d , col=rangi2 ) 
lines( A_seq , mu.mean , lwd=2 ) 
shade( mu.PI , A_seq )
title(ylab =  "                                   Divorce rate",xlab = "                                               Median age marriage")
```

>Divorce rate is also associated with  Marriage rate     m5.2 

```{r}
m5.2 <- quap( 
  alist(
      D ~ dnorm( mu , sigma ) , 
      mu <- a + bM * M ,
      a ~ dnorm( 0 , 0.2 ) , 
      bM ~ dnorm( 0 , 0.5 ) , 
      sigma ~ dexp( 1 )
  ) , data = d )


M_seq <- seq( from=-3 , to=3.2 , length.out=30 ) 
mu <- link( m5.2 , data=list(M=M_seq) )
mu.mean <- apply( mu , 2, mean ) 
mu.PI <- apply( mu , 2 , PI )
# plot it all
plot( D ~ M , data=d , col=rangi2 ) 
lines( M_seq , mu.mean , lwd=2 ) 
shade( mu.PI , M_seq )
title(ylab =  "                                   Divorce rate",xlab = "                                               Marriage rate")
``` 

>DAG Directed Acyclic Graph no.1 

```{r}
library(dagitty)
dag5.1 <- dagitty( "dag{ A -> D; A -> M; M -> D }" )
coordinates(dag5.1) <- list( x=c(A=0,D=1,M=2) , y=c(A=0,D=1,M=0) ) 
drawdag( dag5.1 )
```

>DAG Directed Acyclic Graph  no.2

```{r}
dag5.2 <- dagitty( "dag{ A -> D; A -> M}" )
coordinates(dag5.2) <- list( x=c(A=0,D=1,M=2) , y=c(A=0,D=1,M=0) ) 
drawdag( dag5.2 )
```

>correlation no.1 and no,2 (the same)

```{r}
cor(d$A,d$D)
cor(d$A,d$M)
cor(d$M,d$D)
```

>impliedConditionalIndependencies
 no.2 DAG    D _||_ M | A

```{r}
DMA_dag2 <- dagitty('dag{ D <- A -> M }') 
impliedConditionalIndependencies( DMA_dag2 )
```

>impliedConditionalIndependencies
 no.1 DAG     no conditional independencies, so there is no output to display.

```{r}
DMA_dag1 <- dagitty('dag{ D <- A -> M -> D }') 
impliedConditionalIndependencies( DMA_dag1 )
```

>Divorce rate is associated with median age at marriage & Marriage rate   m5.3

```{r}
m5.3 <- quap( 
  alist(
        D ~ dnorm( mu , sigma ) , 
        mu <- a + bM*M + bA*A , 
        a ~ dnorm( 0 , 0.2 ) , 
        bM ~ dnorm( 0 , 0.5 ) , 
        bA ~ dnorm( 0 , 0.5 ) , 
        sigma ~ dexp( 1 )
      ) , data = d ) 
precis( m5.3 )
```

>focusing just on the slope parameters βA and βM :

```{r}
plot( coeftab(m5.1,m5.2,m5.3), par=c("bA","bM") )
```



```{r}
m5.4 <- quap( 
alist(
M ~ dnorm( mu , sigma ) , 
mu <- a + bAM * A ,
a ~ dnorm( 0 , 0.2 ) , 
bAM ~ dnorm( 0 , 0.5 ) , 
sigma ~ dexp( 1 )
) , data = d )
```

```{r}
mu <- link(m5.4)
mu_mean <- apply( mu , 2 , mean ) 
mu_resid <- d$M - mu_mean
```



>(2) posterior prediction plot
produce a simple posterior predictive check in the divorce example
  begin by simulating predictions, averaging over the posterior.

```{r}
# call link without specifying new data 
# so it uses original data
mu <- link( m5.3 )
# summarize samples across cases 
mu_mean <- apply( mu , 2 , mean ) 
mu_PI <- apply( mu , 2 , PI )
# simulate observations
# again no new data, so uses original data 
D_sim <- sim( m5.3 , n=1e4 )
D_PI <- apply( D_sim , 2 , PI )
```

```{r}
plot( mu_mean ~ d$D , col=rangi2 , ylim=range(mu_PI) , 
xlab="Observed divorce" , ylab="Predicted divorce" )
abline( a=0 , b=1 , lty=2 )
for ( i in 1:nrow(d) ) lines( rep(d$D[i],2) , mu_PI[,i] , col=rangi2 )
```



>(3) Counterfactual plots

>step(1) pick "A" as intervention variable.
 
```{r}
m5.3_A <- quap(
  alist(
     ## A -> D <- M
        D ~ dnorm( mu , sigma ) , 
        mu <- a + bM*M + bA*A , 
        a ~ dnorm( 0 , 0.2 ) , 
        bM ~ dnorm( 0 , 0.5 ) , 
        bA ~ dnorm( 0 , 0.5 ) , 
        sigma ~ dexp( 1 ),
     ## A -> M
        M ~ dnorm( mu_M , sigma_M ), 
        mu_M <- aM + bAM*A,
        aM ~ dnorm( 0 , 0.2 ), 
        bAM ~ dnorm( 0 , 0.5 ), 
        sigma_M ~ dexp( 1 )
     ) , data = d )

```

>step(2)   Define the range of values to set the intervention variable to.

```{r}
A_seq <- seq( from=-2 , to=2 , length.out=30 )
```
>step(3)use the causal model to simulate the values of other variables(excpet A), including the outcome.

```{r}
# prep data
sim_dat <- data.frame( A=A_seq ) 
# simulate M and then D, using A_seq
s <- sim( m5.3_A , data=sim_dat , vars=c("M","D") )
```

> total counterfactual effect of A on D & M, 图

```{r}
plot( sim_dat$A , colMeans(s$D) , ylim=c(-2,2) , type="l" , 
xlab="manipulated A" , ylab="counterfactual D" )
shade( apply(s$D,2,PI) , sim_dat$A )
mtext( "Total counterfactual effect of A on D" )

plot( sim_dat$A , colMeans(s$M) , ylim=c(-2,2) , type="l" , 
xlab="manipulated A" , ylab="counterfactual M" )
shade( apply(s$M,2,PI) , sim_dat$A )
mtext( "Total counterfactual effect of A on M" )
```


#########################################################
#########################################################
##         MILK 
#########################################################
#########################################################

>data

```{r}
library(rethinking) 
data(milk)
d <- milk 

d$K <- standardize( d$kcal.per.g ) 
d$N <- standardize( d$neocortex.perc ) 
d$M <- standardize( log(d$mass) )
```

>model with error ( a lot of NA)

```{r}
m5.5_draft <- quap( 
alist(
K ~ dnorm( mu , sigma ) , 
mu <- a + bN*N ,
a ~ dnorm( 0 , 1 ) , 
bN ~ dnorm( 0 , 1 ) , 
sigma ~ dexp( 1 )
) , data=d )
## there us a lot of NA in d$neocortex.perc

```

>dealing with NA,droping any cases with missing value(NA)

```{r}
dcc <- d[ complete.cases(d$K,d$N,d$M) , ]
d$neocortex.perc
print("---------------------------------------------------------------------")
dcc$neocortex.perc
```

>model (without NA) (DRAFT with BAD prior α & βN)

```{r}
m5.5_draft <- quap( 
                     alist(
                           K ~ dnorm( mu , sigma ) , 
                           mu <- a + bN*N ,
                           a ~ dnorm( 0 , 1 ) , 
                           bN ~ dnorm( 0 , 1 ) , 
                           sigma ~ dexp( 1 )
                   ) , data=dcc )

prior <- extract.prior( m5.5_draft ) 
xseq <- c(-2,2)
mu <- link( m5.5_draft , post=prior , data=list(N=xseq) )
plot( NULL , xlim=xseq , ylim=xseq )
for ( i in 1:50 ) lines( xseq , mu[i,] , col=col.alpha("black",0.3) )
```

>model (without NA) (DRAFT with better prior α & βN )

```{r}
m5.5 <- quap( 
alist(
K ~ dnorm( mu , sigma ) , 
mu <- a + bN*N ,
a ~ dnorm( 0 , 0.2 ) , 
bN ~ dnorm( 0 , 0.5) , 
sigma ~ dexp( 1 )
) , data=dcc )

prior <- extract.prior( m5.5) 
xseq <- c(-2,2)
mu <- link( m5.5 , post=prior , data=list(N=xseq) )
plot( NULL , xlim=xseq , ylim=xseq )
for ( i in 1:50 ) lines( xseq , mu[i,] , col=col.alpha("black",0.3) )
```

>posteriors:
>plot K-N

```{r}
precis( m5.5 )

xseq <- seq( from=min(dcc$N)-0.15 , to=max(dcc$N)+0.15 , length.out=30 ) 
mu <- link( m5.5 , data=list(N=xseq) )
mu_mean <- apply(mu,2,mean) 
mu_PI <- apply(mu,2,PI)
plot( K ~ N , data=dcc ) 
lines( xseq , mu_mean , lwd=2 ) 
shade( mu_PI , xseq )
```
>plot K-M

```{r}
m5.6 <- quap( 
alist(
      K ~ dnorm( mu , sigma ) , 
      mu <- a + bM*M ,
      a ~ dnorm( 0 , 0.2 ) , 
      bM ~ dnorm( 0 , 0.5 ) , 
      sigma ~ dexp( 1 )
      ) , data=dcc ) 
xseq <- seq( from=min(dcc$M)-0.15 , to=max(dcc$M)+0.15 , length.out=30 ) 
mu <- link( m5.6 , data=list(M=xseq) )
mu_mean <- apply(mu,2,mean) 
mu_PI <- apply(mu,2,PI)
plot( K ~ M , data=dcc ) 
lines( xseq , mu_mean , lwd=2 ) 
shade( mu_PI , xseq )
```



>µi   =  α + βN Ni  + βM Mi
>plot    K-N(M=0);   K-M(N=0)

```{r}
m5.7 <- quap( 
alist(
K ~ dnorm( mu , sigma ) , 
mu <- a + bN*N + bM*M , 
a ~ dnorm( 0 , 0.2 ) , 
bN ~ dnorm( 0 , 0.5 ) , 
bM ~ dnorm( 0 , 0.5 ) , 
sigma ~ dexp( 1 )
) , data=dcc ) 
precis(m5.7)

xseq <- seq( from=min(dcc$M)-0.15 , to=max(dcc$M)+0.15 , length.out=30 ) 
mu <- link( m5.7 , data=data.frame( M=0 , N=xseq ) )
mu_mean <- apply(mu,2,mean) 
mu_PI <- apply(mu,2,PI)
plot( NULL , xlim=range(dcc$M) , ylim=range(dcc$K) ) 
lines( xseq , mu_mean , lwd=2 )
shade( mu_PI , xseq )

xseq <- seq( from=min(dcc$M)-0.15 , to=max(dcc$M)+0.15 , length.out=30 ) 
mu <- link( m5.7 , data=data.frame( M=xseq , N=0 ) )
mu_mean <- apply(mu,2,mean) 
mu_PI <- apply(mu,2,PI)
plot( NULL , xlim=range(dcc$M) , ylim=range(dcc$K) ) 
lines( xseq , mu_mean , lwd=2 )
shade( mu_PI , xseq )

```

>coefficient tabel

```{r}
plot( coeftab( m5.5 , m5.6 , m5.7 ) , pars=c("bM","bN") )
```

```{r}

pairs( ~K + M + N , dcc )
```







```{r}
# M -> K <- N 
# M -> N
n <- 100
M <- rnorm( n )
N <- rnorm( n , M ) 
K <- rnorm( n , N - M )
d_sim <- data.frame(K=K,N=N,M=M)
```

```{r}
# M -> K <- N 
# N -> M
n <- 100
N <- rnorm( n )
M <- rnorm( n , N ) 
K <- rnorm( n , N - M )
d_sim2 <- data.frame(K=K,N=N,M=M) 
```

```{r}
# M -> K <- N
# M <- U -> N 
n <- 100
U <- rnorm( n )
N <- rnorm( n , U ) 
M <- rnorm( n , U ) 
K <- rnorm( n , N - M )
d_sim3 <- data.frame(K=K,N=N,M=M)
```

```{r}
dag5.7 <- dagitty( "dag{
M -> K <- N 
M -> N }" )
coordinates(dag5.7) <- list( x=c(M=0,K=1,N=2) , y=c(M=0.5,K=1,N=0.5) ) 
MElist <- equivalentDAGs(dag5.7)
```

```{r}
 drawdag(MElist)
```

>clade

```{r}
levels(d$clade)
d$clade_id <- as.integer( d$clade )
d$clade_id 
```


```{r}
d$K <- standardize( d$kcal.per.g ) 
m5.9 <- quap(
alist(
K ~ dnorm( mu , sigma ), 
mu <- a[clade_id],
a[clade_id] ~ dnorm( 0 , 0.5 ), 
sigma ~ dexp( 1 )
) , data=d )
labels <- paste( "a[" , 1:4 , "]:" , levels(d$clade) , sep="" ) 
plot( precis( m5.9 , depth=2 , pars="a" ) , labels=labels ,
xlab="expected kcal (std)" )
```

```{r}
set.seed(63)
d$house <- sample( rep(1:4,each=8) , size=nrow(d) )
d$house 
```

```{r}
m5.10 <- quap( 
alist(
K ~ dnorm( mu , sigma ),
mu <- a[clade_id] + h[house], 
a[clade_id] ~ dnorm( 0 , 0.5 ), 
h[house] ~ dnorm( 0 , 0.5 ), 
sigma ~ dexp( 1 )
) , data=d )
```
























